Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-31 curvature.tex
consistent enumerate punctation, closes #15

To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by successively adding trees to the graph and efficiently identifying their neighbors:
\begin{enumerate}[label={\arabic*}.]
	\item Let $G$ be an empty graph
	\item For each of the $m$ trees in a specified but arbitrary order,
		\begin{enumerate}
			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index
			\item For each of the $\OhOf{n^2}$ neighbors of $T$
				\begin{enumerate}
					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index
				\end{enumerate}
		\end{enumerate}
\end{enumerate}

@@

We consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
However, some of these moves will result in the same neighboring tree as other rSPR moves.
In particular, where we call the edge connecting the subtree rooted at that node to the rest of the tree the ``node's edge'', we have:
\begin{enumerate}
\item Moving $R$ to its sibling edge results in the same tree, not a neighboring tree,
\item Moving $R$ to its parent edge results in the same tree,
\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge,
\item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
\end{enumerate}
We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive.
We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
Call the portion of the neighborhood of a tree $T$ that can be obtained by moving a subtree $R$ rooted at a node $u$, and that is assigned to $u$, $N(T,u)$.
In this manner, we achieve an alternative solution for computing the degree of a tree in the rSPR graph:

@@

\begin{enumerate}[label={\arabic*}.]
	\item	compute the degree of $T$, $\degree{T}$.
	\item pick a random integer $r$ in the range $[1,\degree{T}]$
	\item label each node $u$ of $T$ by its preorder number and compute the number of nodes in the subtree rooted at each $u$.
\item for each tree node $u$ and while $r > 0$
	\begin{enumerate}
		\item decrease $r$ by $\size{N(T,u)}$
		\item if $r < 0$, let $S$ be the $\abs{r}$ member of $N(T,u)$ and terminate the for loop
	\end{enumerate}
\item return the neighbor $S$
\end{enumerate}

@@

We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
\begin{lemma}
	\label{lem:degree_max_delta_adjacent}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $.
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	We first prove (i).
	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) then maximally balancing the terms in the product $ka$ as follows.

@@

It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
Moreover, nearly all of the neighbors of these pairs approach each either.
We thus conjecture the following:
\begin{conjecture}
	Let $k_n$ be the maximum curvature of the uniform random walk on trees of $n$ leaves.
	Then:
	\begin{enumerate}
		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$
	\end{enumerate}
\end{conjecture}
Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
Next, we bound the minimum curvature of two adjacent trees.

@@

\begin{lemma}
\label{lem:asymptotic}
	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$, and
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
		\item	$\ric{\MH; T,S} = \curvature{\MH; T,S}$, if $\dspr{T,S} > 1$, and
		\item	$\curvature{\MH; T,S} \le \ric{\MH; T,S} \le \curvature{\MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{lemma}

>>>>

To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by successively adding trees to the graph and efficiently identifying their neighbors:
\begin{enumerate}[label={\arabic*}.]
	\item Let $G$ be an empty graph.
	\item For each of the $m$ trees in a specified but arbitrary order:
		\begin{enumerate}
			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index.
			\item For each of the $\OhOf{n^2}$ neighbors of $T$:
				\begin{enumerate}
					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index.
				\end{enumerate}
		\end{enumerate}
\end{enumerate}

@@

We consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
However, some of these moves will result in the same neighboring tree as other rSPR moves.
In particular, where we call the edge connecting the subtree rooted at that node to the rest of the tree the ``node's edge'', we have:
\begin{enumerate}
\item Moving $R$ to its sibling edge results in the same tree, not a neighboring tree,
\item Moving $R$ to its parent edge results in the same tree,
\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge, and
\item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
\end{enumerate}
We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive.
We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
Call the portion of the neighborhood of a tree $T$ that can be obtained by moving a subtree $R$ rooted at a node $u$, and that is assigned to $u$, $N(T,u)$.
In this manner, we achieve an alternative solution for computing the degree of a tree in the rSPR graph:

@@

\begin{enumerate}[label={\arabic*}.]
	\item	Compute the degree of $T$, $\degree{T}$.
	\item Pick a random integer $r$ in the range $[1,\degree{T}]$.
	\item Label each node $u$ of $T$ by its preorder number and compute the number of nodes in the subtree rooted at each $u$.
\item For each tree node $u$ and while $r > 0$:
	\begin{enumerate}
		\item Decrease $r$ by $\size{N(T,u)}$.
		\item If $r < 0$, let $S$ be the $\abs{r}$ member of $N(T,u)$ and terminate the for loop.
	\end{enumerate}
\item Return the neighbor $S$.
\end{enumerate}

@@

We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
\begin{lemma}
	\label{lem:degree_max_delta_adjacent}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	We first prove (i).
	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) then maximally balancing the terms in the product $ka$ as follows.

@@

It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
Moreover, nearly all of the neighbors of these pairs approach each either.
We thus conjecture the following:
\begin{conjecture}
	Let $k_n$ be the maximum curvature of the uniform random walk on trees of $n$ leaves.
	Then:
	\begin{enumerate}
		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
	\end{enumerate}
\end{conjecture}
Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
Next, we bound the minimum curvature of two adjacent trees.

@@

\begin{lemma}
\label{lem:asymptotic}
	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$,
		\item	$\ric{\MH; T,S} = \curvature{\MH; T,S}$, if $\dspr{T,S} > 1$, and
		\item	$\curvature{\MH; T,S} \le \ric{\MH; T,S} \le \curvature{\MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{lemma}
